---
title: "Building an Optimal LEGO Portfolio Under a Budget Constraint"
subtitle: "How much does indivisibility cost a collectibles investor?"
author: "Jeffrey Lee"
date: today
format:
  html:
    theme: cosmo
    toc: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(scales)

overall_index <- readRDS("data/processed/overall_index.rds")
sub_indices <- readRDS("data/processed/sub_indices.rds")
correlation_matrix <- readRDS("data/processed/correlation_matrix.rds")
returns <- readRDS("data/processed/returns.rds")
frontier <- readRDS("data/processed/frontier.rds")
portfolio_2k <- readRDS("data/processed/portfolio_2k.rds")
portfolio_10k <- readRDS("data/processed/portfolio_10k.rds")
portfolio_50k <- readRDS("data/processed/portfolio_50k.rds")

risk_summary <- returns |>
  group_by(themeGroup) |>
  summarize(
    mean_return = mean(return, na.rm = TRUE),
    volatility = sd(return, na.rm = TRUE)
  )
```

## The verdict

Given a fixed budget, this project asks: what does an optimal LEGO aftermarket portfolio actually look like, and how much of the classic portfolio-theory toolkit survives contact with an illiquid, indivisible, thin-data collectible market?

Three findings, in order of importance:

1.  **Nominal price stability masked a real decline.** LEGO's złoty aftermarket price rose steadily from 2018 to 2023, but after adjusting for Polish inflation, real purchasing-power returns peaked in late 2021 and ended below the 2018 starting point by June 2023.
2.  **LEGO theme groups do not diversify each other as well as expected.** Cross-theme correlations of 0.73 to 0.95 contradict the naive assumption that a "collectibles" asset class is internally diversified. Most themes move together, likely riding the same shared macro and market conditions.
3.  **A naive return-maximizing optimizer, ignoring risk, concentrates 100% of any budget into a single theme (Technical), at \$2K, \$10K, and \$50K alike.** A properly risk-adjusted mean-variance frontier, using the real covariance structure, achieves a lower-volatility portfolio than any single theme held alone, by diversifying across four themes instead of one.

The honest headline: naive optimization on this asset class is actively misleading. Rigor changes the answer.

## Data and method

The analysis uses `r nrow(readRDS("data/processed/panel.rds"))` cleaned set-month price observations across `r n_distinct(readRDS("data/processed/panel.rds")$setnum)` LEGO sets, June 2018 through June 2023, sourced from an open academic dataset (Oczkoś, Podgórski, Szczepańska & Boiński, 2024, *Data in Brief* 52:110056, CC BY 4.0), combining aftermarket transaction data (promoklocki.pl) with Brickset set characteristics.

A matched-set hedonic regression (R's `fixest`) estimates monthly price indices while controlling for pieces, minifigure count, and theme, absorbing a separate fixed effect per month to avoid composition-drift bias from an unbalanced panel (793 sets tracked in June 2018, rising to 2,558 by mid-2021, falling to 974 by June 2023). Prices are deflated using Poland's monthly CPI (FRED series POLCPALTT01IXNBM). Portfolio construction compares a mixed-integer program (`ompr`/GLPK) against a continuous quadratic mean-variance frontier (`CVXR`), both using the same underlying covariance structure.

## The price index: nominal vs. real

```{r index-chart}
overall_index |>
  mutate(Date = as.Date(paste0(Date, "-01"))) |>
  ggplot(aes(x = Date)) +
  geom_line(aes(y = nominal_index, color = "Nominal"), linewidth = 1) +
  geom_line(aes(y = real_index, color = "Real (CPI-deflated)"), linewidth = 1, linetype = "dashed") +
  geom_hline(yintercept = 100, linetype = "dotted", color = "gray50") +
  scale_color_manual(values = c("Nominal" = "#2a78d6", "Real (CPI-deflated)" = "#008300")) +
  labs(
    title = "LEGO aftermarket index: nominal price growth outpaced real returns",
    subtitle = "June 2018 = 100. Real index deflated by Polish CPI.",
    x = NULL, y = "Index (June 2018 = 100)", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

By June 2023, the nominal index stood at `r round(tail(overall_index$nominal_index, 1), 1)` while the real, inflation-adjusted index had fallen to `r round(tail(overall_index$real_index, 1), 1)`, below its June 2018 starting value. The gap between the two lines is entirely attributable to Polish inflation, which accelerated sharply through 2022.

## Do LEGO theme groups diversify each other?

```{r correlation-heatmap}
cm <- correlation_matrix
cm_df <- as.data.frame(as.table(cm))
names(cm_df) <- c("theme1", "theme2", "correlation")

ggplot(cm_df, aes(x = theme1, y = theme2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "#2a78d6", mid = "white", high = "#e34948", midpoint = 0.5, limits = c(0,1)) +
  labs(title = "Cross-theme correlation: LEGO themes move together more than expected",
       x = NULL, y = NULL, fill = "Correlation") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")
```

Every pairwise correlation between the seven major theme groups falls between 0.73 and 0.95, all strongly positive. This contradicts the intra-asset-class diversification thesis common in the collectibles-as-investment literature (e.g., Masset & Weisskopf, 2010, on wine): rather than moving independently, LEGO theme groups appear to share substantial common variation, likely driven by shared macroeconomic conditions and overall market sentiment absorbed into the same monthly fixed effects.

## Risk and return by theme

```{r risk-return-table}
risk_summary |>
  arrange(desc(mean_return)) |>
  mutate(
    mean_return = percent(mean_return, accuracy = 0.01),
    volatility = percent(volatility, accuracy = 0.1)
  ) |>
  knitr::kable(col.names = c("Theme", "Mean monthly return", "Monthly volatility"))
```

Technical is both the highest-return and highest-risk theme; Licensed is the steadiest. This spread is what makes the diversification question meaningful, and what the correlation matrix above shows isn't being exploited by simple allocation rules.

## The efficient frontier: what indivisibility costs you

```{r frontier-chart}
ggplot(frontier, aes(x = achieved_risk, y = target_return)) +
  geom_line(color = "#2a78d6", linewidth = 1) +
  geom_point(color = "#2a78d6", size = 2) +
  geom_point(
    data = tibble(risk = 0.138, ret = 0.0142, label = "100% Technical\n(naive integer result,\nall budgets)"),
    aes(x = risk, y = ret), color = "#eb6834", size = 3
  ) +
  geom_text(
    data = tibble(risk = 0.138, ret = 0.0142, label = "100% Technical\n(naive result, all budgets)"),
    aes(x = risk, y = ret, label = label), hjust = -0.05, size = 3, color = "#eb6834"
  ) +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "The efficient frontier vs. the naive integer-optimal result",
    subtitle = "Continuous mean-variance frontier (blue) vs. actual $2K/$10K/$50K portfolios (orange)",
    x = "Monthly risk (volatility)", y = "Monthly expected return"
  ) +
  theme_minimal(base_size = 12)
```

The naive integer-constrained optimizer, used at all three budget levels, sits at the extreme right edge of the frontier: full Technical concentration, `r percent(0.0142, accuracy=0.01)` return at `r percent(0.138, accuracy=0.1)` risk. The frontier shows the same 0.0142 return is only achievable this way, but far better risk/return tradeoffs exist elsewhere on the curve. At the frontier's minimum-risk point, achievable volatility drops to roughly `r percent(min(frontier$achieved_risk), accuracy=0.1)`, meaningfully below Technical's 13.8% and below every other single theme's own volatility.

## Recommended portfolios by budget

::: panel-tabset
### \$2,000

```{r portfolio-2k}
portfolio_2k |>
  mutate(Price = dollar(Price), mean_return = percent(mean_return, accuracy = 0.01)) |>
  knitr::kable(col.names = c("Set", "Theme", "Price", "Expected monthly return"))
```

### \$10,000

```{r portfolio-10k}
portfolio_10k |>
  mutate(Price = dollar(Price), mean_return = percent(mean_return, accuracy = 0.01)) |>
  knitr::kable(col.names = c("Set", "Theme", "Price", "Expected monthly return"))
```

### \$50,000

```{r portfolio-50k}
portfolio_50k |>
  mutate(Price = dollar(Price), mean_return = percent(mean_return, accuracy = 0.01)) |>
  knitr::kable(col.names = c("Set", "Theme", "Price", "Expected monthly return"))
```
:::

These are the outputs of the naive return-maximizing integer program, shown deliberately unmodified: at every budget tested, the optimizer concentrates entirely in Technical sets. This is not a data artifact. It is the mathematically correct behavior of a linear objective with a single budget constraint, sometimes called the "knapsack" property: without a risk term, the optimal solution is always to fill the budget with the single best per-dollar asset. A tested linear risk penalty (return minus a volatility term) failed to change this outcome at any reasonable penalty weight, confirming that true diversification requires the quadratic covariance term used in the frontier above, not a simple per-asset penalty.

## Honest limitations

- **Illiquidity correction was investigated, not applied.** The standard Getmansky-Lo-Makarov unsmoothing correction assumes positive return autocorrelation from appraisal-style smoothing. This data instead showed negative autocorrelation (-0.09 to -0.35 across themes), consistent with month-to-month regression estimation noise from thin trading in some theme-months, rather than classic illiquidity smoothing. Applying the standard correction would have been methodologically inappropriate; raw volatility is reported instead, with this caveat.
- **Individual sets inherit their theme's expected return**, since 61 months of data cannot support reliable per-set return estimates for 6,072 individual sets. This is a real approximation, not a limitation unique to this analysis. It mirrors how thin-trading asset classes are typically handled in practice.
- **Prices are the most recently observed price per set (up to June 2023)**, not live market prices. This is historical, not real-time, portfolio guidance.
- **The frontier is built at the theme level (7 assets)**, not the individual-set level (6,072 assets), because a reliable 6,072x6,072 covariance matrix is not estimable from 61 months of data. The integer-constrained portfolios operate at the set level; the frontier operates at the theme level. This is why the two are compared rather than unified into a single unconstrained-vs-constrained model.
- **Correlations are measured on regression-smoothed indices**, which may understate true month-to-month volatility even where the GLM correction doesn't cleanly apply. This is a genuine, acknowledged tension in this analysis.

## Data and code

Full analysis code, raw data, and reproducibility instructions (via `renv`) are available at [github.com/Watersingman50/lego-portfolio-optimization](https://github.com/Watersingman50/lego-portfolio-optimization).
